---
title: Digging into NJOAG Law Enforcement Officer Diversity data - Part III
author: Tor
date: '2022-07-25'
slug: digging-into-njoag-law-enforcement-officer-diversity-data-part-iii
categories: []
tags:
  - NJOAGLEOD
  - NJOAGUOF
  - R
  - tidymodels
  - xgboost
subtitle: Does law enforcement officer diversity affect use of force?
output:
  blogdown::html_page:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(kableExtra.html.bsTable = TRUE)
options(digits = 2)
renv::use(
  "tor-gu/njoaguof@v1.4.0",
  "tor-gu/njoagleod@v1.1.3",
  "tidyverse@1.3.1",
  "xgboost@1.6.0.1"
)
```

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(kableExtra)
kbl <- function(tbl, caption) {
  knitr::kable(tbl,
               caption = caption,
               format = "html",
               table.attr = "class=\"kable\"") %>%
    kable_styling(bootstrap_options = c("striped", "hover"))
}
```

# Introduction

This is Part III of a series exploring the Law Enforcement Officer Diversity dataset released by the NJ OAG.

- In [Part I](/2022/06/25/digging-into-njoag-law-enforcement-officer-diversity-data-part-i/), we looked at municipal police departments and observed that towns on the Jersey Shore -- especially smaller towns -- had larger police departments relative to population than the rest of the state.
- In [Part II](/2022/06/30/digging-into-njoag-law-enforcement-officer-diversity-data-part-ii/), we started gathering data from other sources to create a combined table of municipal PDs, including use of force data from the [njoaguof](https://github.com/tor-gu/njoaguof/blob/main/README.md) package, plus population, density, income and race and ethinicity data from the US Census. We concluded by observing that agency size is a better predictor than population for use of force incidents.

In this post, we will look at what factors affect the number of use of of force incidents per officer.  In particular, we want to know:

> **Is there a relationship between *law enforcement officer diversity* and the *per-officer use of force rate*?**

Taken at face value, the answer to this question is *yes*.  For example, across municipal agencies, the 2021 per-officer use-of-force rate $F$ is positively correlated with $r_\text{officer black}$, the proportion of officers that are Black ($p=.026$).  On the other hand, the use of force rate is also positively correlated with $r_\text{black}$, the proportion of *residents*  that are Black ($p=7 \times 10^{-6}$).  Of course, $F$, $r_\text{black}$ and $r_\text{officer black}$ are all mutually correlated, and in a joint linear model, $r_\text{black}$ is positively correlated with $F$ while $r_\text{officer black}$ is *negatively* correlated with the use of force rate.

So, refining the question a bit, we want to ask:

> **Across municipal police departments, after accounting for municipal data, is there a relationship between *law enforcement officer diversity* and the *per-officer use of force rate*?**

The perhaps somewhat surprising answer is, for the most part, **no**.  More precisely, we find that, after accounting for municipal household median income, racial demographics, and a regional factor (North Jersey vs South Jersey), there is no significant correlation between the 2021 per-officer rate of use of force on the one hand and officer median age, officer gender ratio, and the ratio of white, Black or Hispanic officers on the other. It is possible that there is a relationship between the ratio of Asian officers and the use of force rate, but the evidence is weak.

But, as it turns out, the problem of finding a connection between police diversity and use of force is not so easy. Evidence in the existing literature for such a relationship is frequently described as "mixed". Here is a sampling:

- [Smith (2003)](https://onlinelibrary.wiley.com/doi/epdf/10.1111/1541-0072.t01-1-00009):  "The findings show that more diversified departments do not have significantly lower levels of police-caused homicides."
- [Hickman and Piquero (2008)](https://journals.sagepub.com/doi/abs/10.1177/0011128708316977): "Minority representation was unrelated to [use of force] complaint rates nor to the percentage of complaints sustained."
- [Smith and Holmes (2014)](https://academic.oup.com/socpro/article-abstract/61/1/83/1626052): "A negative relationship between the ratio of black officers to black citizens and sustained [use of force] complaints exists, indicating that departments more representative of the black community have fewer sustained complaints."
- [Willits and Nowacki (2013)](https://www.tandfonline.com/doi/full/10.1080/10439463.2013.784314): On the connection between representative policing and deadly force, "minority representation ... is also statistically significant and in the expected direction."

Efforts to find such effects are complicated by the fact that agency diversity is probably not exogenous to the use of force rate, as well as a lack of attention (or access) to detailed policing records.  

A significant recent study that avoids these problems is [Ba, Knox, Mummolo and Rivera (2021)](https://www.science.org/doi/10.1126/science.abd8694), which, analyzing millions of detailed records from the city of Chicago, finds that "relative to white officers, Black and Hispanic officers make far fewer stops and arrests, and they use force less often, especially against Black civilians".

This post, far less rigorous than the above cited publications, decidedly does *not* avoid these problems, and we do not claim that there is no connection 
between agency diversity and use of force. Our observation here is simply that, if there is a relationship between municipal police department diversity and the per-officer use-of-force rate in New Jersey, it is not visible in the Law Enforcement Officer Diversity and Use of Force data so far released by the NJ OAG.


## Our approach
This will be another R-centric post. We will start by building models that predict the per-officer use-of-force rate for each municipal police department in terms of the *municipal* data for the community the department serves, including:

  - Population and density
  - Race and ethnicity
  - Household Median Income
  - Regional factors (North Jersey/South Jersey, Jersey Shore)
  
We will privilege interpretable mathematical models over opaque machine learning models, though we will introduce some ML models to see if they are able to turn up relationships the that mathematical models missed.

After we have a model for the use of force rate in terms of the *municipal* data, we see if the *agency* data has any additional predictive value. The agency data
includes:

  - Officer male/female ratio
  - Officer mean age
  - Officer race and ethnicity

Rather than jumping to our final models, we will explore several alternatives, though we will not recapitulate all the tuning and feature pruning. We will be using the [tidymodels](https://www.tidymodels.org/) idiom for building recipes and doing cross validation.

## Packages used in this post

In this post, I'll be using the following packages.

```{r echo=FALSE, message=FALSE}
pkgkbl <- function(packages) {
  packages %>% map(packageDescription) %>%
    map(~ c(.$Package, .$Title, .$Version)) %>%
    as_tibble(.name_repair = "minimal") %>% t() %>% 
    as_tibble(.name_repair = "unique") %>%
    rename(Package = ...1, Title = ...2, Version = ...3) %>%
    kbl(caption = NULL)
}
c("njoagleod", "njoaguof", "gridExtra", "sf", "tidyverse", "tidymodels", "tidycensus", "tigris",  "vip", "xgboost") %>%  pkgkbl()
```

This is not a tutorial for using these packages, but all the R code used to generate the plots, tables and models will be included.

The OAG packages can be installed from github:

```{r eval=FALSE}
install_github("tor-gu/njoaguof")
install_github("tor-gu/njoagleod")
```

# Preparing the training data
```{r echo=FALSE}
load(here::here("static", "data", "use_of_force_data.Rda"))
```
In [Part II](/2022/06/30/digging-into-njoag-law-enforcement-officer-diversity-data-part-ii/) we assembled a `combined_table` of municipal agencies including most of the data we will need. A script to recreate this table (along with `nj_municipality_map`) may be found [here](https://gist.github.com/tor-gu/c03164e55d851b2e31c87ea5b5b6591b). Before using this data, we need to do some additional preparation. We need to:

- Add a column for `incident_rate_per_officer`. This will be the outcome value in our models.
- Add a column for `officer_years`, the product of `officer_count` and `partial_year`. This will be the case weight in our models.
- Filter out the rows with `NA` values for `incident_rate_per_officer`.
- Add a field for North Jersey.
- Arrange the columns so the agency and municipal predictors are last.
- Split the data into test and training sets.

For our purposes, "North Jersey" means the municipalities in the 11 northern-most counties in the state:

```{r}
# Northern counties
northern_nj_counties <- c("Bergen County", "Essex County", "Hudson County",
  "Hunterdon County", "Middlesex County", "Morris County", "Passaic County",
  "Somerset County", "Sussex County", "Union County", "Warren County")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Map of northern counties
library(tigris)
options(tigris_use_cache = TRUE)

counties(state = "NJ", class = "sf") %>%
  mutate(northern_nj = NAMELSAD %in% northern_nj_counties) %>%
  ggplot(aes(geometry = geometry)) +
  geom_sf(aes(fill = northern_nj), show.legend = FALSE) +
  geom_label(mapping = aes(label = NAME),
             stat = "sf_coordinates",
             size = 3) +
  labs(title = "Eleven northern NJ counties") +
  theme(
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.background = element_rect(fill = "lightblue"),
    panel.grid.major = element_line(color = "lightblue")
  )
```
There are [other definitions](https://en.wikipedia.org/wiki/North_Jersey) of the North Jersey cultural region. Our definition is on the expansive side, including the three northernmost border counties, Hunterdon, Somerset and Middlesex, and comprises about 62% of the population.

Here is how we construct our new combined table, `ct`. Note that we mark
`officer_years` as type `tidymodels::importance_weights`.
```{r message=FALSE}
library(tidymodels)
library(njoagleod)
# Build a combined table:
# - Add incident_rate_per_officer (output)
# - Add officer_years (weight)
# - Add northern_nj
# - put all predictors at the end.
ct <- combined_table %>%
  mutate(incident_rate_per_officer = incident_rate_est / officer_count,
         officer_years = importance_weights(officer_count * partial_year)
  ) %>%
  filter(!is.na(incident_rate_per_officer)) %>%
  left_join(municipality, by = "GEOID") %>%
  mutate(northern_nj = case_when(
    county %in% northern_nj_counties ~ TRUE,
    TRUE ~ FALSE
  )) %>% 
  select(-county, -municipality) %>%
  relocate(GEOID, incident_rate_per_officer, 
           officer_count, incident_count, partial_year, officer_years,
           incident_rate_est)
```
```{r echo=FALSE, message=FALSE}
ct %>% select(GEOID, incident_rate_per_officer, 
           officer_count, officer_years, officer_mean_age, officer_r_male) %>% head(5) %>% kbl("ct, selected columns")
```

For the purpose of model building, we split this into training and testing sets.

```{r}
# split into training and testing
set.seed(37)
ct_split <- initial_split(ct, strata = "incident_rate_per_officer", prop = .75)
ct_train <- training(ct_split)
ct_test <- testing(ct_split)
```
Our training set has 295 rows. 100 rows have been reserved as test data.

## A note on our model outcome variable, `incident_rate_per_officer`
Before we begin building the models, let's pause a moment to consider our 
model output variable, `incident_rate_per_officer`, which is  `incident_rate_est` divided by `officer_count`.  The value `incident_rate_est` represents our estimate for the annual rate of use of force incidents per year for each agency.^[Based on 2021 use of force incidents. For departments with a full year of data, `incident_rate_est` is just `incident_count`. For agencies with only a partial year of data, we divide by `partial_year`, which is itself an estimate. See [Part II](/2022/06/30/digging-into-njoag-law-enforcement-officer-diversity-data-part-ii/) for details.] We use the per-officer rate rather than the per-resident rate, because `incident_rate_est` is more strongly correlated with `officer_count` than population.  But is `incident_rate_per_officer` really independent of `officer_count`?

In our training data, the two value *are* correlated, but only at $p=.11$
```{r message=FALSE, warning=FALSE}
ct_train %>% ggplot(aes(y = incident_rate_per_officer, x = officer_count)) +
  geom_point(aes(alpha = partial_year)) +
  geom_smooth(aes(weight = partial_year), color = "lightyellow") +
  geom_smooth(aes(weight = partial_year), method = "lm", se = FALSE) +
  scale_x_log10() + scale_y_log10() +
  labs(title = "Incident rate not quite independant of officer count",
       subtitle = "(Training data only)") +
  theme(legend.position = "bottom")
```
```{r}
model <- lm(incident_rate_per_officer ~ officer_count, 
   ct_train, weights=partial_year) %>% tidy()
```
```{r echo=FALSE}
options(digits = 4)
model %>% kbl("Incident rate ~ officer count")
```

We will proceed as if `incident_rate_per_officer` is independent of `officer_count`, but it is something to keep an eye on.

# Building the models
As we mentioned above, our approach is:

- First, build a model of `incident_rate_per_officer` in terms of the *municipal* predictors, and excluding the *agency* predictors.
- Next, try to model the residuals of the municipal model with the *agency* predictors.

## Municipal models
Our available predictors are `population`, `density`, `r_white`, `r_black`, 
`r_native_american`, `r_asian`, `r_pacific_islander`, `r_other_races`, `r_two_or_more_races`, `r_non_hispanic_or_latino`, `r_hispanic_or_latino`,
`household_median_income`, `shore_town` and `northern_nj`.

We will begin with linear models.

### Linear municipal model
With a linear model, we will find that the best predictors are `household_median_income` and `northern_nj`, along with some combination of the
racial demographics. We get better results when the the race predictors are transformed -- we use $log(1+t)$ here, though $1/(1+t)$ also works well.

Below, we construct a linear model with `r_white`, `r_black`, `r_asian` and `r_other_races`. We use the `tidymodels` idiom for building up our models in
terms of recipes. Our model fit to the training data is `fit_lm`.

```{r}
predictors_lm <-
  c(
    "household_median_income",
    "northern_nj",
    "r_white",
    "r_black",
    "r_asian",
    "r_other_races"
  )

mod_lm <- linear_reg()

recipe_lm <- recipe(incident_rate_per_officer ~ ., data = ct_train) %>% 
  step_rm(-incident_rate_per_officer,         # Outcome
          -officer_years,                     # Case weights
          -all_of(predictors_lm)) %>%         # Predictors
  step_log(starts_with("r_"), offset = 1)

wf_lm <- workflow() %>%
  add_model(mod_lm) %>%
  add_recipe(recipe_lm) %>%
  add_case_weights(officer_years)

fit_lm <- wf_lm %>%
  fit(data=ct_train)
```
All of the predictors are significant in this model, but it is worth noting that the adjusted $R^2$ only reaches $.325$.  (We will not do much better than this in any of our models.)
```{r echo=FALSE}
options(digits=7)
```
```{r}
summary(fit_lm %>% extract_fit_engine())$adj.r.squared
tlm <- fit_lm %>% tidy()
```
```{r echo=FALSE}
tlm %>% kbl("Linear model fitted on training data")
```
The largest contributions are coming from `household_median_income` and `northern_nj`: There are lower rates of use of force in North Jersey and in higher income towns. In this model, the difference between North Jersey and South Jersey is about the same as a difference of $72,000 in household median income.
Interpreting the coefficients of the racial demographic predictors is less straightforward. We will return to them in a subsequent post.

### A 'linear-plus' model incorporating the Jersey Shore and officer count
The `shore_town` flag does not appear in our linear model, and adding it does not improve the model. However, recalling the interaction we saw in Part I between `shore_town` and `officer_count` -- as well as our suspicion about the relationship between `officer_count` and `incident_rate_per_officer` -- we may be
tempted to add the interaction `shore_town * officer_count` to our model. Indeed, we do see an improvement, though we get our best model if we remove `officer_count`.  We call this the 'linear-plus' model.

```{r}
predictors_lm_plus <- c(predictors_lm, "shore_town", "shore_officer_count")
recipe_lm_plus <- recipe(incident_rate_per_officer ~ ., data = ct_train) %>% 
  step_mutate(shore_officer_count = shore_town * officer_count) %>%
  step_rm(-incident_rate_per_officer,         # Outcome
          -officer_years,                     # Case weights
          -all_of(predictors_lm_plus)) %>%    # Predictors
  step_log(starts_with("r_"), offset = 1)

wf_lm_plus <- wf_lm %>%
  update_recipe(recipe_lm_plus)

fit_lm_plus <- wf_lm_plus %>%
  fit(data = ct_train)

summary(fit_lm_plus %>% extract_fit_engine())$adj.r.squared
tlm <- fit_lm_plus %>% tidy()
```
```{r echo=FALSE}
tlm %>% kbl("'Linear-plus' model, fitted on training data")
```
In the linear-plus model, `lm_plus`, the interaction is captured by `shore_officer_count`, which is just the `officer_count` when `shore_town` is TRUE, and is `0` when `shore_town` is false.  So the `lm_plus` model has dependence on the officer count for shore towns only.

This model has a marginally better $R^2 = .35$, but at the cost of adding
a predictor, `officer_count`, that we did not want to add.

### Cross validation

Let's check the robustness of our models by using cross validation.

```{r}

#### CV of linear models
set.seed(37)
folds <- vfold_cv(ct_train, v = 10, repeats = 5)
metrics_lm <- wf_lm %>%
  fit_resamples(folds) %>% 
  collect_metrics()
metrics_lm_plus <- wf_lm_plus %>%
  fit_resamples(folds) %>% 
  collect_metrics()

mt <- lst(metrics_lm, metrics_lm_plus) %>%
  bind_rows(.id = "model") %>%
  mutate(model = str_sub(model, 9)) %>%
  select(-.config)
```
```{r echo=FALSE}
options(digits=3)
mt %>% kbl("Cross validation metrics for the linear and linear-plus models")
```

We observe that on cross validated data, the linear model slightly beats the linear-plus model on both RMSE (.424 vs .436) and $R^2$ (.254 vs .233). The advantage the linear-plus model showed before we did cross validation was probably
a bit of overfitting.  For now, we will take `lm` as our best model.

### Problems with the linear municipal models 
One obvious problem with the linear models is that they may predict negative incident rates. In fact, we do have one negative prediction in the training data.
```{r warning=FALSE}
fit_lm %>%
  extract_fit_engine() %>%
  augment(ct_train) %>% 
  ggplot(aes(x = .fitted, y = incident_rate_per_officer)) +
  geom_point(aes(size = officer_count, alpha = partial_year)) +
  geom_abline(color = "red") +
  ylim(0, 2.5) +
  theme(legend.position = "bottom") +
  labs(title = "Linear model fit",
       subtitle = "(Training data only)")
```
A second problem is that the model may be under-predicting the higher rates, though it is hard to say from this graph because of theheteroskedasticity:
```{r}
par(mfrow = c(1, 2))
plot(fit_lm %>% extract_fit_engine(), which = c(1, 3))
```


### Log-linear municipal model
We will try to address the linear model problems with a log-linear model,
$\log y = \alpha + \sum\beta_ix_i + \epsilon$.  
```{r}
# Log-linear model
mod_llm <- linear_reg() %>%
  set_engine("glm", family = Gamma("log"))
```
We can reuse our linear model recipe, except that we will have to filter out
the rows with no incidents. We note that we can expect this to cause a bias away from zero for small departments.
```{r}
recipe_llm <- recipe_lm %>% 
  step_filter(incident_rate_per_officer > 0)
wf_llm <- wf_lm %>%
  update_model(mod_llm) %>%
  update_recipe(recipe_llm)
fit_llm <- wf_llm %>%
  fit(ct_train)
```
We do not have negative predictions or heteroskedasticity with the log-linear model.
```{r warning=FALSE}
fit_llm %>%
  extract_fit_engine() %>%
  augment(ct_train %>% filter(incident_rate_per_officer > 0),
          type.predict = "response") %>% 
  ggplot(aes(x = .fitted, y = incident_rate_per_officer)) +
  geom_point(aes(size = officer_count, alpha = partial_year)) +
  geom_abline(color = "red") +
  ylim(0, 2.5) + xlim(0, 1.5) +
  theme(legend.position = "bottom") +
  labs(title = "Log-Linear model fit",
       subtitle = "(Training data only)")
```
```{r}
par(mfrow = c(1, 2))
plot(fit_llm %>% extract_fit_engine(), which=c(1, 3))
```

Let's also try a log-linear-plus variant, similar to linear-plus:
```{r}
recipe_llm_plus <- recipe_lm_plus %>% 
  step_filter(incident_rate_per_officer > 0)
wf_llm_plus <- wf_llm %>%
  update_recipe(recipe_llm_plus)
fit_llm_plus <- wf_llm_plus %>% 
  fit(ct_train)
```

### Cross validation
Let's get cross validation metrics for the two log-linear models and compare to the linear models.
```{r}
metrics_llm <- wf_llm %>%
  fit_resamples(folds) %>% 
  collect_metrics()
metrics_llm_plus <- wf_llm_plus %>%
  fit_resamples(folds) %>% 
  collect_metrics()
mt <- lst(metrics_lm, metrics_llm, metrics_lm_plus, metrics_llm_plus) %>%
  bind_rows(.id = "model") %>%
  mutate(model = str_sub(model, 9)) %>%
  select(-.config)
```
```{r echo=FALSE}
options(digits=3)
mt %>% kbl("Cross validation metrics for the linear and log-linear models")
```

The differences here are very slight, but the log-linear model `llm` performs at least as well as the linear model `lm`, while adding some desirable features (no negative predictions, homoskedasticity). The `lm_plus` and `llm_plus` are not adding anything that would justify the compromise we made in constructing them (adding `shore_officer_count`).

So our best model so far is the log-linear model `llm`.

### ML municipal model
Before turning to the agency predictors, let's see if machine learning algorithms can turn up any patterns that our linear models are missing.

We are not interested in replacing our interpretable linear model with an opaque
ML model that performs only marginally better, so we will not spend a huge 
amount of time here tuning. We will use the [xgboost](https://xgboost.readthedocs.io/en/stable/) package to build our model.^[Before settling on xgboost, we used [h2o AutoML](https://docs.h2o.ai/h2o/latest-stable/h2o-docs/automl.html) to do an initial search. The top performing models tended to be GBMs or stacked models. They did not perform as well as our best xgboost tune.]

We will begin by constructing a recipe for xgboost. We let xgboost use all the 
municipal predictors except `r_pacific_islander` and `r_native_american`, which are usually very close to zero.
```{r}
counts <- ct_train %>%
  count(r_native_american > 0.01, r_pacific_islander > 0.005)
```
```{r echo=FALSE}
counts %>% kbl("Very few towns in the training set with significant Native American or Pacific Islander populations")
```
We will do the same `log` transform of the `r_` fields as we did in our linear models. Note we have to convert the logical columns (`shore_town` and `northern_nj`) to integers.
```{r}
municipal_predictors <- ct %>% select(population:northern_nj) %>% names()
recipe_xgb <- recipe(incident_rate_per_officer ~ ., data = ct_train) %>%
  step_rm(-incident_rate_per_officer, -officer_years,
          -all_of(municipal_predictors)) %>%
  step_rm(r_pacific_islander, r_native_american) %>%
  step_log(starts_with("r_"), offset = 1) %>%
  step_integer(northern_nj, shore_town)
```
Here is the best tune we found for this recipe:
```{r}
# XGBoost model with tuned parameters
mod_xgb <- boost_tree(
  mtry = .48,
  trees = 600,            # nrounds
  min_n = 24,             # min_child_weight
  tree_depth = 2,         # max_depth
  learn_rate = .0185,     # eta
  loss_reduction = .001,  # gamma
  sample_size = .715,
  stop_iter = 50          # early_stopping_rounds
) %>% 
  set_engine("xgboost",
             alpha = .1,
             lambda = 17500,
             nthread = 12,
             counts = FALSE) %>%
  set_mode("regression")  
# XGBoost workflow
wf_xgb <- workflow() %>%
  add_model(mod_xgb) %>%
  add_recipe(recipe_xgb) %>%
  add_case_weights(officer_years)
```
### Comparison with log-linear model

If we look at the cross validation metrics and compare the XGBoost model to our existing models, we see that we have a small but noticeable improvement in both $R^2$ and RMSE.
```{r}
metrics_xgb <- wf_xgb %>%
  fit_resamples(folds) %>% 
  collect_metrics()
mt <- lst(metrics_lm, 
    metrics_llm, 
    metrics_lm_plus, 
    metrics_llm_plus,
    metrics_xgb) %>%
  bind_rows(.id = "model") %>%
  mutate(model = str_sub(model, 9)) %>%
  select(-.config)
```
```{r echo=FALSE}
mt %>% kbl("Cross validation metrics for the linear, log-linear, and XGBoost models")
```
Next, let's compare the shape of the predictions generated by the log-linear and ML models:
```{r warning=FALSE}
fit_xgb <- wf_xgb %>%
  fit(ct_train)
ct_train_aug <- bind_cols(
  ct_train,
  fit_llm %>% predict(ct_train) %>% rename(llm.pred = .pred),
  fit_xgb %>% predict(ct_train) %>% rename(xgb.pred = .pred)
)
ct_train_aug %>%
  pivot_longer(llm.pred:xgb.pred, names_to = "model", values_to = "pred") %>%
  ggplot(aes(x = incident_rate_per_officer, y = pred)) +
  geom_point(aes(size = officer_count, alpha = partial_year)) +
  geom_abline() +
  xlim(0,2.1) +
  facet_wrap("model") +
  labs(title = "Log-linear and XGBoost predictions by incident rate",
        subtitle = "(Training data only)")
ct_train_aug %>%
  ggplot(aes(y = llm.pred, x = xgb.pred)) +
  geom_point(aes(size = officer_count, alpha = partial_year)) +
  geom_abline() +
  xlim(.3, 1) +
  labs(title = "Log-linear predictions by XGBoost predictions",
        subtitle = "(Training data only)")
```
Here we see that that the ML model is far more conservative than the log linear 
model, never making predictions outside of the band $[.35,1]$, though the two models are pretty tightly correlated.


Next, let's find out what the XGBoost model is seeing that the linear models are missing, by first taking a look at variable importance.
```{r message=FALSE}
library(vip)
library(gridExtra)
grid.arrange(
  vip(fit_llm %>% extract_fit_engine()) + 
    labs(title = "Log-linear model vip"),
  vip(fit_xgb %>% extract_fit_engine()) + 
    labs(title = "XGBoost model vip"),
  ncol = 2
)
```
The ML model didn't find much use for the variables omitted from our linear models, but it does rate `r_asian` more highly. Let's see if we can find out where `fit_xgb` is making better predictions than `fit_llm` by breaking the training
set down by the top three variables, `r_asian`, `household_median_income`, and `northern_nj`:
```{r}
tbl <- ct_train_aug %>%
  mutate(region = if_else(northern_nj, "NORTH", "SOUTH"),
         income = if_else(household_median_income < 89000, "LOW", "HIGH"),
         r_asian = case_when(
           r_asian < .023 ~ "LOW",
           r_asian < .064 ~ "MEDIUM",
           TRUE ~ "HIGH"
         )) %>%
  group_by(r_asian, region, income) %>%
  summarize(count = n(),
            rate.mean = weighted.mean(incident_rate_per_officer,
                                      as.double(officer_years)),
            llm.mean = weighted.mean(abs(llm.pred), 
                                     as.double(officer_years)),
            xgb.mean = weighted.mean(abs(xgb.pred), 
                                     as.double(officer_years)),
            llm.mae = weighted.mean(abs(llm.pred-incident_rate_per_officer), 
                                    as.double(officer_years)),
            xgb.mae = weighted.mean(abs(xgb.pred-incident_rate_per_officer), 
                                    as.double(officer_years)),
            .groups="drop"
  ) %>%
  arrange(xgb.mae - llm.mae)
```
```{r echo=FALSE}
tbl %>% kbl("Mean Average Error of Log-linear and XGBoost model across several categories")
```
```{r message=FALSE, warning=FALSE}
labeller = as_labeller(
  c("LOW" = "Low Asian density", "MEDIUM" = "Medium Asian density", 
    "HIGH" = "High Asian density",
    "NORTH" = "North Jersey", "SOUTH" = "South Jersey")
)
ct_train_aug %>%
  mutate(region = if_else(northern_nj, "NORTH", "SOUTH"),
         r_asian = factor(case_when(
           r_asian < .023 ~ "LOW",
           r_asian < .064 ~ "MEDIUM",
           TRUE ~ "HIGH"
         )),
         r_asian = fct_relevel(r_asian, "LOW", "MEDIUM", "HIGH"),
         xgb_v_llm =
           abs(xgb.pred - incident_rate_per_officer) -
           abs(llm.pred - incident_rate_per_officer)) %>%
  ggplot(aes(x = household_median_income, y = incident_rate_per_officer,
             color = xgb_v_llm)) +
  geom_point(aes(size = officer_count, alpha = partial_year)) +
  geom_smooth(aes(weight = as.double(officer_years)), method = "lm", se = FALSE,
              color = "grey") +
  annotate("rect", xmin = 0, xmax = 250000, ymin = .37, ymax = 1, 
           alpha = .1, fill = "green") +
  scale_color_gradient2(low = "red", mid = "white", high = "blue",
                        limits = c(-.3, .3), oob = squish) +
  scale_x_continuous(labels = label_number(suffix = "K", scale = .001)) +
  ylim(0, 2.2) +
  facet_grid(rows = vars(region), cols = vars(r_asian), labeller = labeller) +
  theme(legend.position = "none") +
  labs(title = "Model performance by Asian population density, region and income",
       subtitle = "Red = ML model better, Blue = Log-linear model better")
```

(The green bar represents the prediction range for the ML model)

General conclusions that we can draw from this table and graph:

- Household income becomes less important as the Asian population density increases.
- The log-linear model performs best relative to the ML model when the Asian population density is low.
- The log-linear model does better with high incident rates in South Jersey and low rates in North Jersey.
- Conversely, the ML model does better with low incident rates rates in South Jersey and high incident rates in North Jersey.

The South Jersey, low-income, low-Asian-density sector is particularly 
interesting:

  - It has the highest mean rate of use of force per officer (1.12).
  - It has the highest mean prediction for both ML (1.008) and log-linear models (.912).
  - It is where the log-linear model (.345 MAE) has its best performance relative to ML model (.387 MAE).
  - It has the highest MAE for the ML model in an absolute sense.
  - It contains 6 of the 10 cases where the ML model outperforms the log-linear model.
  - It contains 6 of the 10 cases where the log-linear model outperforms the ML model.

We could plausibly gain a marginal improvement in the linear models by adding some
interaction terms between these three variables, but we will not pursue this here.

## Agency models
Let us now see what can be gained by considering the *agency* predictors.
Our approach will be to take the residuals from the log-linear model from the previous section, and try to build a model for the residuals in terms of the
agency predictors.

### Linear agency models
Let's start very simply with a linear model. First we set up the table
with the residuals added.
```{r}
ct_train_agcy <-
  bind_cols(
    ct_train,
    wf_llm %>% fit(ct_train) %>% predict(ct_train) %>% rename(llm.pred = .pred),
  ) %>%
  mutate(llm.resid = incident_rate_per_officer - llm.pred)

```
Now we set up a linear model of the residual in terms of the agency predictors.
```{r}
mod_agcy_lm <- linear_reg()
agency_predictors <- ct_train_agcy %>%
  select(officer_mean_age:officer_r_race_na) %>% names()
recipe_agcy_lm <- recipe(llm.resid ~ ., data = ct_train_agcy) %>% 
  step_rm(-llm.resid,                         # Outcome
          -officer_years,                     # Case weights
          -all_of(agency_predictors))         # Predictors

wf_agcy_lm <- workflow() %>%
  add_model(mod_agcy_lm) %>%
  add_recipe(recipe_agcy_lm) %>%
  add_case_weights(officer_years)

fit_agcy_lm <- wf_agcy_lm %>%
  fit(ct_train_agcy)

tbl <- fit_agcy_lm %>% tidy()
```
```{r echo=FALSE}
tbl %>% kbl("Log Linear residuals ~ agency predictors")
```
The model is not promising. It has $p=.423$ and adjusted $R^2=.0003$. Log-transforming the `officer_r_` fields, as we did for the `r_` fields in the municipal models, does not turn out to be helpful. If we iteratively remove the worst predictor and refit, we arrive at our best linear model, which has a single predictor, `officer_r_asian`.
```{r}
recipe_agcy_lm_asian <- recipe(llm.resid ~ officer_r_asian + officer_years,
                               data = ct_train_agcy)
wf_agcy_lm_asian <- wf_agcy_lm %>%
  update_recipe(recipe_agcy_lm_asian)
fit_agcy_lm_asian <- wf_agcy_lm_asian %>%
  fit(ct_train_agcy)
tbl <- fit_agcy_lm_asian %>% tidy()
```
```{r echo=FALSE}
tbl %>% kbl("Log Linear residuals ~ officer_r_asian")
```
This is not a very strong model, but on cross validation it looks worse, as it
is beaten by the empty model (RMSE .418 vs .417 for the empty model).  Visually, it is hard to see how the model could be much improved by non-linear terms without overfitting.
```{r message=FALSE, warning=FALSE}
ct_train_agcy %>%
  ggplot(aes(x = officer_r_asian, y = llm.resid)) +
  geom_point(aes(size = officer_count, alpha = partial_year)) +
  geom_smooth(aes(weight = as.double(officer_years)), color = "lightyellow") +
  scale_x_log10() +
  theme(legend.position = "bottom") +
  ylim(-1, 1.8) + 
  labs(title = "Log Linear residuals by officer_r_asian",
       subtitle = "(Training data only)")
```
Let's now see if we can find an ML algorithm that will turn up anything that we missed.

### ML agency models
We will use XGBoost again. Here is our model and best tune:
```{r}
recipe_agcy_xgb <-
  recipe(llm.resid ~ ., data = ct_train_agcy) %>%
  step_rm(GEOID:partial_year, incident_rate_est, population:llm.pred)

mod_agcy_xgb <- boost_tree(
  trees = 3000,           # nrounds
  min_n = 14,             # min_child_weight
  tree_depth = 11,        # max_depth
  learn_rate = .008,      # eta
  loss_reduction = 44,    # gamma
  sample_size = 1,
  stop_iter = 50          # early_stopping_rounds
) %>% 
  set_engine("xgboost",
             alpha = .003,
             lambda = .003,
             colsample_bytree = 1,
             nthread = 12,
             counts = FALSE) %>%
  set_mode("regression")

wf_agcy_xgb <- workflow() %>%
  add_model(mod_agcy_xgb) %>%
  add_recipe(recipe_agcy_xgb) %>%
  add_case_weights(officer_years)
```
Unfortunately this model barely outperforms the empty model (RMSE .416 vs .417 for the empty model, $R^2$ = .03).  The top variables in this model are `officer_r_asian`, which we examined when evaluating linear models, and `officer_mean_age`.  
```{r}
fit_agcy_xgb <- wf_agcy_xgb %>%
  fit(ct_train_agcy)
fit_agcy_xgb %>%
  extract_fit_engine() %>% vip()
```
Here is now `fit_agcy_xgb` uses `officer_mean_age`.  
```{r warning=FALSE}
augment(fit_agcy_xgb, ct_train_agcy) %>%
  pivot_longer(c(".pred", "llm.resid"), 
               names_to = "residual_type", 
               values_to = "residual") %>%
  ggplot(aes(x = officer_mean_age, y = residual, color = residual_type)) +
  geom_point() +
  scale_color_hue(labels = c("Predicted", "Actual")) +
  ylim(-1.1, 2) +
  guides(color = guide_legend(title = "Residual Type")) +
  theme(legend.position = "bottom") +
  labs(title = "ML model predictions vs officer_mean_age")
```
Possibly there is something going on with younger agencies, but the prediction
spike in the narrow band around age 36 appears to be an overfit driven
by a few extreme values.

In any case, the ML model is barely distinguishable from the empty model.

## Final fit
Having selected the log-linear model, let's perform a final fit and see 
how it performs against the test data.
```{r}
last_fit_llm <- wf_llm %>%
  last_fit(ct_split)
tbl <- last_fit_llm %>%
  collect_metrics()
```
```{r echo=FALSE}
tbl %>% kbl("Log-linear model applied to test data")
```
The RMSE (.395) is a little better than the cross validated RMSE on the training set, but the $R^2$ (.217) is a bit worse.

# Conclusions

- Using a log-linear model of the use-of-force rate for municipal agencies, we can account for a bit above 20% of the variance in terms of household median income, several racial demographic factors, and a North Jersey regional flag.
- After accounting for the municipal factors with the log-linear model, essentially no additional variance can be explained by agency diversity factors.

# Next
In the next post, we will look at outliers and interpretations of the log-linear municipal model.
